---
title: "Practical machine learning course Project"
author: "Jacob Mokgadi"
date: "18/02/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Synopsis

In this project, we used data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants to predict the manner in which they did the exercise. This is the “classe” variable in the training set, We train 2 models: Decision Trees and Random Forests. The training and testing data was provided and it was determined that Random Forests was the more accurate prediction method.

## Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: (http://groupware.les.inf.puc-rio.br/har) (see the section on the Weight Lifting Exercise Dataset).

The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

## Loading data and Libraries

Loading all necessary libraries

```{r}
library(rpart)
library(caret)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(randomForest)
set.seed(12345)
```

Getting the data

```{r}
traindat <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testdat <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
```

Loading the data

```{r}
training <- read.csv(url(traindat), na.strings=c("NA","#DIV/0!",""),stringsAsFactors = TRUE)
testing <- read.csv(url(testdat), na.strings=c("NA","#DIV/0!",""),stringsAsFactors = TRUE)
```

## Partitioning the training set into to
The training data will be partitioned in to 60% training and 40% testing

```{r}
inTrain <- createDataPartition(y=training$classe, p=0.6, list=FALSE)
myTraining <- training[inTrain, ]; myTesting <- training[-inTrain, ]
dim(myTraining); dim(myTesting)
```

## Cleaning the data
Cleanind the data for zero or missing values, we run this code to view NZV (non-zero variables)

```{r}
myDataNZV <- nearZeroVar(myTraining, saveMetrics=TRUE)
```

Create another subset without NZV variables

```{r}
myNZVvars <- names(myTraining) %in% c("new_window", "kurtosis_roll_belt", "kurtosis_picth_belt",
"kurtosis_yaw_belt", "skewness_roll_belt", "skewness_roll_belt.1", "skewness_yaw_belt",
"max_yaw_belt", "min_yaw_belt", "amplitude_yaw_belt", "avg_roll_arm", "stddev_roll_arm",
"var_roll_arm", "avg_pitch_arm", "stddev_pitch_arm", "var_pitch_arm", "avg_yaw_arm",
"stddev_yaw_arm", "var_yaw_arm", "kurtosis_roll_arm", "kurtosis_picth_arm",
"kurtosis_yaw_arm", "skewness_roll_arm", "skewness_pitch_arm", "skewness_yaw_arm",
"max_roll_arm", "min_roll_arm", "min_pitch_arm", "amplitude_roll_arm", "amplitude_pitch_arm",
"kurtosis_roll_dumbbell", "kurtosis_picth_dumbbell", "kurtosis_yaw_dumbbell", "skewness_roll_dumbbell",
"skewness_pitch_dumbbell", "skewness_yaw_dumbbell", "max_yaw_dumbbell", "min_yaw_dumbbell",
"amplitude_yaw_dumbbell", "kurtosis_roll_forearm", "kurtosis_picth_forearm", "kurtosis_yaw_forearm",
"skewness_roll_forearm", "skewness_pitch_forearm", "skewness_yaw_forearm", "max_roll_forearm",
"max_yaw_forearm", "min_roll_forearm", "min_yaw_forearm", "amplitude_roll_forearm",
"amplitude_yaw_forearm", "avg_roll_forearm", "stddev_roll_forearm", "var_roll_forearm",
"avg_pitch_forearm", "stddev_pitch_forearm", "var_pitch_forearm", "avg_yaw_forearm",
"stddev_yaw_forearm", "var_yaw_forearm")
myTraining <- myTraining[!myNZVvars]
```
Chech the new N of observations 

```{r}
dim(myTraining)
```
Then we take out the first column dataset

```{r}
myTraining <- myTraining[c(-1)]
```

Then we clean all variables with too many NAs.

```{r}
trainingV3 <- myTraining #creating another subset to iterate in loop
for(i in 1:length(myTraining)) { #for every column in the training dataset
        if( sum( is.na( myTraining[, i] ) ) /nrow(myTraining) >= .6 ) { #if n?? NAs > 60% of total observations
        for(j in 1:length(trainingV3)) {
            if( length( grep(names(myTraining[i]), names(trainingV3)[j]) ) ==1)  { #if the columns are the same:
                trainingV3 <- trainingV3[ , -j] #Remove that column
            }   
        } 
    }
}
#check the new N?? of observations
dim(trainingV3)
```

set back to our set

```{r}
myTraining <- trainingV3
rm(trainingV3)
```

We do the same cleaning exercise for the testing data

```{r}
clean1 <- colnames(myTraining)
clean2 <- colnames(myTraining[, -58]) #already with classe column removed
myTesting <- myTesting[clean1]
testing <- testing[clean2]
dim(myTesting)
```
```{r}
dim(testing)
```
We need the coerce the data into the same type, for proper functioning of our analysis methods

```{r}
for (i in 1:length(testing) ) {
        for(j in 1:length(myTraining)) {
        if( length( grep(names(myTraining[i]), names(testing)[j]) ) ==1)  {
            class(testing[j]) <- class(myTraining[i])
        }      
    }      
}
#And to make sure Coertion really worked, simple smart ass technique:
testing <- rbind(myTraining[2, -58] , testing) #note row 2 does not mean anything, this will be removed right.. now:
testing <- testing[-1,]
```

## Decision trees for prediction

```{r}
modFitA1 <- rpart(classe ~ ., data=myTraining, method="class")
fancyRpartPlot(modFitA1)
```

Prediction and using confusion matrix to test the results
```{r}
predictionsA1 <- predict(modFitA1, myTesting, type = "class")
myTesting$classe<-as.factor(myTesting$classe)

```

Test the results with a confusion matrix
```{r}
confusionMatrix(predictionsA1, myTesting$classe)
```


## Using Random Forests

```{r}
modFitB1 <- randomForest(classe ~. , data=myTraining)
```

Predicting sample error

```{r}
predictionsB1 <- predict(modFitB1, myTesting, type = "class")
predictionsB2 <- predict(modFitB1, testing, type = "class")
```
Testing the results with a confusion matrix

```{r}
confusionMatrix(predictionsB1, myTesting$classe)
```

## Conclusion

Random forests yields better results, as expected. Random Forests give an accuracy of 99.7% as compared to a 87% accuracy for Decision trees






